---
title: "Machine Learning 1"
author: 'Tianru Zhang (PID: A15432834)'
date: "10/21/2021"
output: pdf_document
---

First step is clustering methods
#Kmeans clustering
The function in the base R to do Kmeans clustering is called `Kmeans()`.

First, make up some data which nobody knows what the answer should be.
```{r}
tmp <-c(rnorm(30,-3),rnorm(30,3))
x<-cbind(x=tmp,y=rev(tmp))
plot(x)
```
 
>Q1. Can we do Kmeans() to cluster this data setting k=2 and nstart=20?


```{r}
km<-kmeans(x,centers=2, nstart=20)

km
```

Thus, we can do this clustering.

>Q2. How many points are there in each cluster?

```{r}
km$size
```

>Q3. What component of the result object details cluster assignment/membership?


```{r}
km$cluster
```

>Q4. What component of the result object details the cluster centers?


```{r}
km$centers
```




>Q5 Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
 
 
```{r}
plot(x,col=km$cluster)
points(km$centers,col="blue",pch=16, cex=2)
```
 
 
 #hclust
 
 A big limitation with k-means is that we have to tell it the value of k(the number of clusters we want)
 Analyze with hclust, and demonstrate the use of dist(),hclust(),plot() and cutree() functions to do the clustering.
 Generate dendrograms and return cluster assignments to membership vectors. 
 
 
```{r}

hc<-hclust(dist(x))
hc
```

There is a plot method for hclust result objects.
Let's see it!!

```{r}
plot(hc)
```



To get the clustering membership vector we have to do a bit more work. 
We have to cut the tree dendrogram where we consider making sense. 
Use the `cutree()`function

```{r}
cutree(hc, h=6)
```

You can also call `cutree()` setting k=number of groups/clusters 

```{r}
grps<-cutree(hc, k=2)

```


Make the plot of groups

```{r}
plot(x,col=grps)
```


```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```


#Principal component data analysis(PCA)

##Getting data

>Q1.How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?


```{r}
dim(x)
```

There are 17 rows and 5 columns. Whoops! There should only be four, because the first column is literally the names of rows.


```{r}
head(x)
```


The first way to fix the column name issue:
```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)

```


```{r}
dim(x)
```

The second way to fix:
```{r}
x <- read.csv(url, row.names=1)
head(x)
```
 >Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
 
 I prefer the second method, because with the first approach, if we run x <- x[,-1]), multiple times, it will lose column information and override the table.
 
 
 
##Spotting major differneces and trends
Now that the data looks good, we will analyze it with some conventional plots.

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?


```{r}
barplot(as.matrix(x), beside=FALSE, col=rainbow(nrow(x)))
barplot(as.matrix(x), col=rainbow(nrow(x)))
```
By setting the beside parameter to FALSE, or leaving the argument, will generate the new plot as above.

 The stacked plot is not helpful to look at.
 
 
 
 >Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?
 
 
 
```{r}
pairs(x, col=rainbow(17), pch=16)
```
 This graph plots each country against each other. If a point lies on the diagonal, it means the two countries have the same amount of food consumption for that type of food.
 
 
 
 >Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
 
 The N. Ireland has one food consumption that significantly exceeds any other country in UK. But it's hrad to tell which food it is based on the plot alone.
 
#PCA to the rescue!
 
 The main function in base R for PCA is `prcomp()`.
 This function transposes our data.
 
```{r}
# Use the prcomp() PCA function 
#t(x) takes the transpose of our data
pca <- prcomp( t(x) )
summary(pca)
```
 
```{r}
attributes(pca)
```


```{r}
plot(pca$x[,1],pca$x[,2])
```



>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.



```{r}
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.


```{r}

color<-c("red","blue","purple","orange")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500),col=color)
text(pca$x[,1], pca$x[,2], colnames(x),col=color)
```


```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v

```

```{r}
## or the second row here...
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

#Diving Deeper


```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )

```
This PC1 analysis shows difference in fresh fruit and alcoholic drinks 

>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?
Fresh potatoes and soft drinks prominantely shows difference in consumption.

```{r}
## Lets focus on PC2 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```




#PCA of RNA-seq data
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
>Q10: How many genes and samples are in this data set?

```{r}
dim(rna.data)
```
There are 100 genes and 10 samples. 
