---
title: "Lab10_git"
output: github_document
---
---
title: "Haloween Candy"
author: 'Tianru Zhang (PID: A15432834)'
date: "10/28/2021"
output: github_document
---

```{r}

url<-"https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
#candy_file <- read.csv(url)

candy <- read.csv(url, row.names=1)
head(candy,n=5)
```

>Q1.How many candy types: 85

There are 85 types of candies in total.
```{r}
nrow(candy)
```

>Q2: How many fruity candy types?

there are 38 of them.
```{r}
sum(candy$fruity)
```
#2. Favourite candy


```{r}
candy["Twix", ]$winpercent
```
>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Happy Cola", ]$winpercent
```
My fav is Haribo Happy Cola and its winpercent value is 34%.


>Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat", ]$winpercent
```
The percent is 76.77% for Kit Kat

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
The winpercent is 49.65% for Tootsie Roll Snack Bars.

Side-note: the skimr::skim() function

There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.


```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent, it looks on a very different scale.


>Q7. What do you think a zero and one represent for the candy$chocolate column?

I think it indicates that the candy$chocolate stands for categorical data 


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```